---
title: "Example Solution"
author: "Daniel W. Kennedy"
date: "September 24, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Example

```{r}
devtools::find_rtools(debug = TRUE)
library(rstan)

data("nhtemp")
plot(nhtemp,ylab = expression(Temperature ~(degree~"F")),xlab = "Year")
```

The `nhtemp` data-set in R is a time series of temperatures (measured in $^\circ$F) collected in New Haven, Connecticut over 60 years from 1912-1971. The proposed model for the process is that the temperature depends on the previous year's temperature as well global temperatures, which are assumed to rise linearly each year:
\[
  y_i = \alpha + \beta y_{i-1} +\gamma i + \epsilon_i,
\]
where
\[
  \epsilon_i \sim \text{Normal}{\left(0,\sigma^2\right)}.
\]

Explanation for each parameter is:

  - $\alpha$ is an "intercept" temperature (it does not correspond exactly correspond to the expected temperature at $i=0$ as the stationary mean is $\alpha/\left(1-\beta\right)$$^\circ$F)
  - $\beta$ is the first order autocorrelation parameter. If $\beta$ is 0, then there is no autocorrelation, whereas if $\beta$ is less than or greater than 0, then the current value $y_i$ will be dependent on the previous value $y_{i-1}$.
  - $\gamma$ is the linear trend, where each year the average temperature increases by $\gamma$$^\circ$F.

Create a Stan model and use MCMC sampling in R to:

  1. Estimate the first order autocorrelation in the model.
  2. Estimate the long term linear trend.
  3. Forecast the temperature for the next 20 years (1972 - 1981) after the data-set.



Use this code as a template for your Stan model:

```{stan , output.var = "stan_template", eval = FALSE}
data {
  int<lower=0> N;
  int<lower=2> N_new;
  vector[N] y;
}

parameters {

}

model {
  for (n in 2:N){
    y[n] ~ normal(FILL, FILL);
  }
}

generated quantities{
  real y_new[N_new];
  y_new[1] = normal_rng(FILL, FILL);
  for (n in 2:N_new){
    y_new[n] = normal_rng(FILL, FILL);
  }
}
```

## Solution:

The following Stan model should allow us to estimate the model parameters and forecast temperature for proceeding years:

```{stan , output.var = "full_stan_model", eval = FALSE}
data {
  int<lower=0> N;
  int<lower=2> N_new;
  vector[N] y;
}

parameters {
  real gamma;
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  for (n in 2:N)
    y[n] ~ normal(alpha + beta * y[n-1] + gamma * n, sigma);
}

generated quantities{
  real y_new[N_new];
  y_new[1] = normal_rng(alpha + beta * y[N] + gamma * (N + 1), sigma);
  for (n in 2:N_new){
    y_new[n] = normal_rng(alpha + beta * y_new[n-1] + gamma * (n + N), sigma);
  }
}
```

We now compile the model and sample for 5000 iterations on 4 parallel chains:

```{r}
ts_ar1.stanmodel <- stan_model(file = "models/ts_ar1.stan")

ts_ar1.stanfit <- sampling(ts_ar1.stanmodel,
                           data = list(
                             N = length(nhtemp),
                             N_new = 20,
                             y = as.vector(nhtemp)),
                           iter = 5000,
                           chains = 4,
                           warmup = 1000)
```

We can now use the fabulous `ggmcmc` package to plot our MCMC chains and make sure that they have converged:

```{r}
library(ggmcmc)
ggs_traceplot(ggs(ts_ar1.stanfit),family = c("gamma|alpha|beta|sigma"))
ggs_autocorrelation(ggs(ts_ar1.stanfit),family = c("gamma|alpha|beta|sigma"))
ggs_running(ggs(ts_ar1.stanfit),family = c("gamma|alpha|beta|sigma"))
ggs_density(ggs(ts_ar1.stanfit),family = c("gamma|alpha|beta|sigma"))
```

We can get numerical summaries for your parameters as well:
```{r}
print(ts_ar1.stanfit,pars = c("gamma","alpha","beta","sigma"))
```

We therefore find there is compelling evidence for a linear trend, where temperatures rise by 0.03(95\% CI: 0.01,0.05)$\circ$F per year. There is also very little evidence of autocorrelation, once the linear trend is taken into account, as the credible interval for $\beta$ covers 0 (-0.17,0.37).


We can extract and plot our forecasted temperatures using the `As.mcmc.list()` function. There may be a simpler way of converting the MCMC samples to a data-frame, however using the `do.call()` and `lapply()` functions are used here:

```{r}
predictions <- As.mcmc.list(ts_ar1.stanfit,pars = "y_new")
predictions <- 
  do.call(
    rbind.data.frame, 
    lapply(1:4, function(x){return(cbind(predictions[[x]],chain = x,iter = 1:nrow(predictions[[x]])))}))

library(reshape2)
predictions <- predictions %>% melt(id = c("chain","iter"))
# Convert the year to a numeric variable:
levels(predictions$variable) <- 1971 + 1:length(levels(predictions$variable))
predictions$variable <- as.numeric(as.character(predictions$variable))

# Calculate the mean, upper and lower quartiles, and bounds for a 90% Prediction Interval.
prediction_summary <- 
predictions %>% group_by(variable) %>% 
  summarise(mean = mean(value),
            upper_q = quantile(value,0.75),
            lower_q = quantile(value,0.25),
            upper_ci = quantile(value,0.95),
            lower_ci = quantile(value,0.05))

# Plot the real data:
plot <- ggplot() + geom_line(data = data.frame(y = as.vector(nhtemp), x = 1912:1971),aes(x=x,y=y))

# Plot the mean prediction
plot <- plot + geom_line(data = prediction_summary, mapping = aes(x = variable, y = mean))

# Plot the upper and lower quartiles as a ribbon:
plot <- plot + geom_ribbon(
  data = prediction_summary, 
  mapping = aes(x = variable, ymin = lower_q, ymax = upper_q),alpha = 0.3, fill = "red")

# Plot the 90% Prediction interval as a lighter ribbon.
plot <- plot + geom_ribbon(
  data = prediction_summary, 
  mapping = aes(x = variable, ymin = lower_ci, ymax = upper_ci),alpha = 0.2, fill = "red")
plot <- plot + labs(y = expression(Temperature ~(degree~"F")),x = "Year")
plot
```

